---
title: "Ritchies Rule Plus"
author: "Theodore Van Rooy"
date: "January 6, 2015"
output: html_document
runtime: shiny
---


# To Trade or Invest, that is the question.

### Or... is all you need to succeed proper bet sizing?

There once was a time when I believed that all my mathematical and statistical knowledge would make me a fortune in the markets.  

After 5 years of "chasing the wind" as the writer of Ecclesiastes might say, I retired my trading chops in favor of pursuing lower hanging fruit.


## Ritchies Rule aka bet sizing

Recently though, an acquaintence of mine, [Mark Ritchie](http://markritchie.me), wrote a book on trade sizing in which he demonstrates his formula (similar to the Kelly criterion) for calculating how to much to risk.

The basic motivation of Ritchies Rule is trading the amount of money which keeps you alive to fight another day.

Below is a simplified version of this app which you can fill estimated winning percentage into.  I simplified to show you the likelihoods of winning outcomes and their densities.  Mark has an [app](https://itunes.apple.com/us/app/ritchirule/id410837344?mt=8) on the iTunes store which you can buy for $0.99 that does more than the following if you're interested.

Also, for comparison, I showed similar returns for just putting all of your money into the S&P500 and going long (for any 1000 day stretch in it's history).

```{r, echo=FALSE}
library(shiny)
library(ggplot2)

sp500 = read.csv('data/SP500.csv')
sp500[["pGain"]] = 1+(lag(sp500$Close)-sp500$Close) / lag(sp500$Close)
sp500[is.na(sp500)]=1

inputPanel(
  numericInput('winP',label = 'Winning Percentage (range = 1 to 100)', value = 52, min = 1,max = 100),
  numericInput('betP', label = 'Bet Size as a percentage of total capital', value=8, min=1, max=100),
  sliderInput('avgLossP',label = 'Avg. Trade Loss/Gain size', value=10, min=1, max=100)
)


renderPlot({
  #100 simulated returns color coded by end profit
  withProgress(message = 'Calculating Ritchies Rule',
                 detail = 'This may take a while...', value = 0, {
                   
      profits = data.frame(returns=rep(1, 1000), time=1:1000, col=.5)            
      for (i in 1:200) {
        #incProgress(1/100)
        returns = abs(rnorm(1000, 0, sd = input$avgLossP/100/3))*ifelse(runif(1000)>input$winP/100, -1, 1)
        betSize = input$betP/100*returns+1
        profits = rbind(profits, data.frame(returns=cumprod(betSize), time=1:1000, col=betSize[length(betSize)]))
      }
    })
  
  ggplot(profits, aes(x=time, y=returns))+geom_hex(alpha=.75)+geom_density2d(color="red",size=2)+#(color="grey", alpha=.5)+
    ylim(.4, 1.6)+
    labs(title='Density of % Returns (likelihood) with Ritchies Rule Optimal bet percentage')
  
 
})

renderPlot({
   
  withProgress(message = 'Calculating Long on S&P500 Returns',
                 detail = 'This may take a while...', value = 0, {
                   
      profits = data.frame(returns=rep(1, 1000), time=1:1000, col=.5)            
      for (i in 1:200) {
        #incProgress(1/100)
        startPoint=sample(1:(nrow(sp500)-1001), 1)
        returns = sp500$pGain[startPoint:(startPoint+999)]
        profits = rbind(profits, data.frame(returns=cumprod(returns), time=1:1000, col=.5))
      }
    })
  
  ggplot(profits, aes(x=time, y=returns))+geom_hex(alpha=.75)+geom_density2d(color="red",size=2)+#(color="grey", alpha=.5)+
    ylim(.4, 1.6)+
    labs(title='Density of % Returns (likelihood) going long on S&P500')
  
})
```



The above plot shows us what is most likely to occur using the Ritchie Rule betting process with normally distributed returns (mean=0, 99th quantile = avg. gain/loss size) and the specified winning percentage and average trade movement.

**Observations** 

*  If you can steadily rack up more winning trades with large movements then you can get good risk reduced returns with good bet sizing
*  But if you can't get both good movement (~10% per trade) or your winning percentage isn't greater than 52%, bet sizing simply delays the inevitable
*  Realistically, your chances for achieving good returns with low risk and low stress are probably maximized on the S&P 500

**A conclusion**

Living to fight another day is great, but ultimately your trading has to be worthwhile and with manageable stress.  So unless you can scale your position upwards with limited downside (for instance call options) and achieve significantly more winning than losing trades, then reducing your bet size wont help you beat the S&P 500.

In fact, we can figure out the average time to bust for any given winning percentage, bet size and average size of win or loss for a trade.  

Let's look at what happens with a losing trading system in which 51% of trades lose, and specify that we can gain or lose up to 90% of our trade.  What then is the average time for a system to die with varying bet sizing?

```{r echo=FALSE}

library(dplyr)
profits = data.frame(returns=rep(1, 1000), time=1:1000, trial=1, betSize=0)      
betS = seq(1,100, by=10)
  for(b in betS){
      for (i in 1:10) {
        #incProgress(1/100)
        returns = abs(rnorm(1000, 0, sd = .3))*ifelse(runif(1000)>49/100, -1, 1)
        betSize = b/100*returns+1
        profits = rbind(profits, data.frame(returns=cumprod(betSize), time=1:1000, trial=i, betSize=b))
      }
}

bustByBet = profits %.% 
  group_by(betSize, trial) %.%
  arrange(time) %.%
  summarize(bustTime = ifelse(any(returns<.01), which(returns<.01)[1], 1000)) %.%
  group_by(betSize) %.%
  summarize(meanBust = mean(bustTime))

ggplot(bustByBet, aes(x=betSize, y=meanBust))+geom_line()+
  labs(title="Mean Time to Bust", xtitle='Bet Size %', ytitle='Days')


```


So up to about 12.5% betting size works here to not bust within 1000 days with a slightly losing track record.  

### Note that you never bust 1 time in any 1000 day period while trading the S&P 500

# Does Mark Ritchie's Rule work?

Yes.  But it wont necessarily make you any more money than just playing the S&P 500 long.

All trading is risky and stressful too.

#  Further thoughts on Mark Ritchie's Rule

## Issue 1, comissions and slippages

Trading isn't free.  There are two types of costs:

*  Entry/exit costs commonly known as comissions (usually from $5 to $20 for stocks, options bonds and $20-50 for futures) 
*  Slippage

Slippage is, well, a slippery topic.  Most traders know about it, but many investors wont.  Slippage is the phenomenon that accompanies trading with stops.  Generally one exits a position when it moves in the unprofitable direction.  Selling in a buyers market means that your target "stop" price often "slips" a few additional points in the wrong direction.

## Issue 2, fat tails

Mark Ritchie explains his rule using the law of averges and coin flips.  And the above charts are constructed on normally distributed returns and "geometrically" distributed winning/losing streaks.  Unfortunately the markets are not coin flips, nor are gains normally distributed.  Markets tend to have longer down and up streaks than one would desire.  Furthermore, markets tend to have "black swan" events, like the 2008 housing bubble burst.

## Issue 3, it takes a lot of work to trade

And at the end of the day if your returns aren't much better than the S&P500... why bother doing all that work, taking on risk and stressing out?





