---
title: "Ritchies Rule Plus"
author: "Theodore Van Rooy"
date: "January 6, 2015"
output: html_document
runtime: shiny
---


# To Trade or Invest, that is the question.

There once was a time when I believed that all my mathematical and statistical knowledge would make me a fortune in the markets.  

After 5 years of "chasing the wind" as the writer of Ecclesiastes might say, I retired my trading chops in favor of pursuing lower hanging fruit.

Recently though, an acquaintence of mine, [Mark Ritchie](http://markritchie.me), wrote a book on trade sizing in which he demonstrates his formula (similar to the Kelly criterion) for calculating how to much to risk.

I don't disagree with it, but I also know that there's more to it than what he presents.  

## Ritchies Rule aka bet sizing

The basic motivation of Ritchies Rule is trading the amount of money which keeps you alive to fight another day.

If you're at the casino playing blackjack you don't want to bet all your money on one go (actually you do if you want the best chance of doubling your money, but that's for another analysis).  Ritchies Rule helps you determine the amount that allows you to play as long as possible without bankruptcy.  Why?  Cause you love gambling!

Okay here's the basic rule:

$$ riskAmount = \frac{T}{W}*\frac{\frac{w}{l}+1-\frac{t}{w}}{\frac{p}{w}} $$

Below is a simplified version of this app which you can fill estimated winning percentage into.  I simplified to show you the average trend and the 95th quantile returns with a simple coin flip betting scheme.  Mark has an [app](https://itunes.apple.com/us/app/ritchirule/id410837344?mt=8) on the iTunes store which you can buy for $0.99 that does more than the following if you're interested.

```{r, echo=FALSE}
library(shiny)
library(ggplot2)

inputPanel(
  numericInput('winP',label = 'Winning Percentage (range = 1 to 100)', value = 52, min = 1,max = 100),
  numericInput('betP', label = 'Bet Size as a percentage of total capital', value=8, min=1, max=100)
)


renderPlot({
  #100 simulated returns color coded by end profit
  withProgress(message = 'Calculating Ritchies Rule',
                 detail = 'This may take a while...', value = 0, {
                   
      profits = data.frame(returns=rep(1, 1000), time=1:1000, col=.5)            
      for (i in 1:200) {
        #incProgress(1/100)
        returns = abs(rnorm(1000, 0, sd = .003))*ifelse(runif(1000)>input$winP/100, -1, 1)
        betSize = input$betP/100*returns+1
        profits = rbind(profits, data.frame(returns=cumprod(betSize), time=1:1000, col=betSize[length(betSize)]))
      }
    })
  
  ggplot(profits, aes(x=time, y=returns))+geom_hex(alpha=.75)+geom_density2d(color="red",size=2)+#(color="grey", alpha=.5)+
    labs(title='Density of % Returns (likelihood) with Ritchies Rule Optimal bet percentage')
})

```

The above plot shows us what is most likely to occur using the Ritchie Rule betting process with normally distributed returns and the specified winning percentage.

** Observations ** 

*  Bet size determines how quickly you go broke
*  Lower bet size means lower returns as well

** A conclusion **

Living to fight another day is great, but ultimately your trading has to be worthwhile.  So unless you can scale your position upwards with limited downside (for instance a call option), reducing your bet size wont help you beat the S&P 500.

In fact, we can figure out the average time to bust for any given winning percentage and bet size.  

Here we are going to specify a slightly losing percentage automatically (49% of trades lose) just so that we can see when any system would die (as they all will).

```{r echo=FALSE}

library(dplyr)
profits = data.frame(returns=rep(1, 1000), time=1:1000, trial=1, betSize=0)      
betS = seq(1,100, by=10)
  for(b in betS){
      for (i in 1:10) {
        #incProgress(1/100)
        returns = abs(rnorm(1000, 0, sd = .3))*ifelse(runif(1000)>40/100, -1, 1)
        betSize = b/100*returns+1
        profits = rbind(profits, data.frame(returns=cumprod(betSize), time=1:1000, trial=i, betSize=b))
      }
}

bustByBet = profits %.% 
  group_by(betSize, trial) %.%
  arrange(time) %.%
  summarize(bustTime = ifelse(any(returns<.01), which(returns<.01)[1], 1000)) %.%
  group_by(betSize) %.%
  summarize(mean(bustTime))


```



#  My additions to Mark Ritchies Rule

Okay, great, we live to fight another day!  But what if we just want to make money? There are 3 additional aspects of trading to cover.

And at the end of the day there's one simple decision we have to make, take on high stress and trade actively, or dump money into the S&P 500 and let it ride (while you sip Mai Thais on the beach).

## Issue 1, comissions and slippages

Trading isn't free.  There are two types of costs:

*  Entry/exit costs commonly known as comissions (usually from $5 to $20 for stocks, options bonds and $30-50 for futures) 
*  Slippage

Slippage is, well, a slippery topic.  Most traders know about it, but many investors wont.  Slippage is the phenomenon that accompanies trading with stops.  Generally one exits a position when it moves in the unprofitable direction.  Selling in a buyers market means that your target "stop" price often "slips" a few additional points in the wrong direction.

## Issue 2, fat tails

Mark Ritchie explains his rule using the law of averges and coin flips.  And the above charts are constructed on normally distributed returns and "geometrically" distributed winning/losing streaks.  Unfortunately the markets are not coin flips, nor are gains normally distributed.  Markets tend to have longer down and up streaks than one would desire.  Furthermore, markets tend to have "black swan" events, like the 2008 housing bubble burst.

## Issue 3, if you only want to survive just invest in S&P 500 ETF's

It's actually pretty simple, if your goal is to make a consistent return on investment and never go bust just invest in S&P 500 ETFs.  They have no comissions and low managment funds.  I recommend VTI (Vangaurd Total Return).  

But if you like the rush of entering orders and obsessing over every trend then sure... go for some day trading!

There's only 1 caviat, you must out predict your neighbors consistently.

# Ritchies Rule Plus

Use the app below to look at what happens when you trade on actual data with winning trade percentages and realistic trade comissions vs just going long on the S&P 500.  If you think you can consistently beat the market then go for it!

As for me... I'll continue investing the majority of my savings in S&P500 and the occassional high risk options trade to satisfy my inner gambler :-)




